Assessment of a Shen-Fourier turbulent channel flow solver for large-scale simulations

### 2016-01-29 paper.tex
Triangle t

Equations (\ref{eq:v0}) and (\ref{eq:w0}) are also discretized in time using 
Crank-Nicolson and Adams-Bashforth, and we obtain
\begin{align}
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{v}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{v}^{n}_0 + \D{B}\hat{\mathcal{H}}_{y,0}^{n+1/2}, \label{eq:v00} \\
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{w}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{w}^{n}_0 + \D{B}\hat{\mathcal{H}}_{z,0}^{n+1/2}, \label{eq:w00}
\end{align}
where we have used notation $\hat{v}^{n}_0 = \hat{v}(l, 0, 0, t_n)$, 
$\hat{\mathcal{H}}^{n}_{y,0} = \hat{\mathcal{H}}_y(l, 0, 0, t_n)$ and similar 
for $\hat{w}^n_0$ and $\hat{\mathcal{H}}^n_{z,0}$.

@@

To solve Eq. (\ref{eq:gfin}) we perform an LU decomposition of the coefficient 
matrix $\D{H}$. $\D{H}$ consists of one subdiagonal, and every 
second upper diagonal is zero. As such, we may split the matrix into two 
smaller (and decoupled) matrices for odd and even coefficients $\D{H}^e_{k,j} = 
\D{H}_{2k,2j}$ and $\D{H}^o_{k,j} = \D{H}_{2k+1,2j+1}$, where the two matrices 
$\D{H}^{e,o}$ are of type upper 
Hessenberg, with only 4 distinct values on each row. This structure of $\D{H}$ 
allows us to perform a very efficient tailored LU decomposition, as shown in 
Algorithm (\ref{alg:lu}), and a very efficient tailored solve through 
Algorithm (\ref{alg:lusolve}). The efficiency follows from the fact that all 
$U^{o,e}_{kj} = U^{o,e}_{kk+2} \, \forall \, j > k+2$, i.e., the upper 
tridiagonal matrix has a maximum of three distinct values for each row. 
Regarding Algorithm 
(\ref{alg:lusolve}), we note that the ability to decouple the system into odd 
and even coefficients leads to two subsystem that may be trivially solved 
simultaneously in two threads by 
letting one thread call LUODDEVEN with even coefficients, and the second thread 
call it for odd.

@@

\EndProcedure
\end{algorithmic}

\end{algorithm}

>>>>

Equations (\ref{eq:v0}) and (\ref{eq:w0}) are also discretized in time using 
Crank-Nicolson and Adams-Bashforth, and we obtain
\begin{align}
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{v}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{v}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{y,0}^{n+1/2} , 
\label{eq:v00} \\
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{w}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{w}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{z,0}^{n+1/2}, 
\label{eq:w00}
\end{align}
where we have used notation $\hat{v}^{n}_0 = \hat{v}(l, 0, 0, t_n)$, 
$\hat{\mathcal{H}}^{n}_{y,0} = \hat{\mathcal{H}}_y(l, 0, 0, t_n)$ and similar 
for $\hat{w}^n_0$ and $\hat{\mathcal{H}}^n_{z,0}$.

@@

To solve Eq. (\ref{eq:gfin}) we perform an LU decomposition of the coefficient 
matrix $\D{H}$. $\D{H}$ consists of one subdiagonal, and every 
second upper diagonal is zero. As such, we may split the matrix into two 
smaller (and decoupled) matrices for odd and even coefficients $\D{H}^e_{k,j} = 
\D{H}_{2k,2j}$ and $\D{H}^o_{k,j} = \D{H}_{2k+1,2j+1}$, where the two matrices 
$\D{H}^{e,o}$ are of type upper 
Hessenberg, with only 4 distinct values on each row. This structure of $\D{H}$ 
allows us to perform a very efficient tailored LU decomposition, as shown in 
Algorithm (\ref{alg:lu}), and a very efficient tailored solve through 
Algorithm (\ref{alg:lusolve}). The efficiency follows from the fact that all 
$U^{o,e}_{kj} = U^{o,e}_{kk+2} \, \forall \, j > k+2$, i.e., the upper 
tridiagonal matrix has a maximum of three distinct values for each row. 
Regarding Algorithm 
(\ref{alg:lusolve}), we note that the ability to decouple the system into odd 
and even coefficients leads to two subsystem that may be trivially solved 
simultaneously in two threads by letting one thread call LUSolveoddeven with 
even coefficients, and the second thread call it for odd. For optimal 
performance, the odd and even coefficient would then need to be stored 
contiguously in computer memory, and not alternately, which would be the normal 
way of storing the coefficients.

@@

\EndProcedure
\end{algorithmic}
\end{algorithm}

The biharmonic operator $\N{H}$ is more challenging to solve efficiently, but 
it is still possible to find an algorithm that is $O(N)$ for $N$ unknowns. Let 
first $\N{H}_kj$ be decoupled into odd $\N{H}^o_{k,j}= \N{H}_{2k,2k}$ and even 
$\N{H}^e_{k,j}= \N{H}_{2k+1,2k+1}$ coefficients. Both systems can be solve with 
the same algorithm, so we show only the details for the even coefficients. Note 
that $\N{H}$ is the sum of three matrices $\N{L}, \N{A}$ and $\N{B}$ and can be 
written
\begin{equation}
\N{H} = c_0\N{L} + c_1\N{A} + c_2 \N{B},
\end{equation}
where $c_0, c_1$ and $c_2$ are constants. The matrix $\N{A}$ contains only 
three nonzero diagonals at $k-2, k, k+2$,  whereas matrix $\N{B}$ contains five 
nonzero diagonals at $k-4, k-2, k, k+2, k+4$. The $\N{L}$ matrix is given as
\begin{align}
 \N{L}_{kk} &= -4(k+1)(k+2)^2 \\
 \N{L}_{kj} &= p(k)q(j) + r(k)s(j), \quad j = k+2, k+4, \ldots N_x-4
\end{align}
where 
\begin{align}
p(k) &= 8 k (k+1)(k+2)(k+4)\pi, &q(j) = \frac{1}{j+3} \\
r(k) &= 24(k+1)(k+2)\pi, &s(j) =  \frac{(j+2)^2}{(j+3)}
\end{align} 
For each row in $\N{H}_{kj}$ we have that $\N{H}_{kj}=c_0\N{L}_{kj}\, \forall j 
> k+4$. A direct LU decomposition of $\N{H}^e$ leads to a lower triangluar 
matrix $l^e_{kj}$ with two nonzero diagonals and an upper triangular matrix 
$U^e_{kj}$ that can be written as
\begin{equation}
\N{U}^e_{kj} = a^e(k)q(2j) + b^e(k)s(2j), \quad j > k+2
\end{equation}
where $a^e(k)$ and $b^e(k)$ are two new vectors that can be pre-computed 
recursively as
\begin{align}
a^e[0] &= p(0),\\
b^e[0] &= r(0), \\
a^e[1] &= p(2) - l^e_{10}a[0]\\
b^e[1] &= r(2) - l^e_{10}b[0]\\
a^e(k) &= p(2k) - l^e_{k,k-1} a^e(k-1) - l^e_{k,k-2}a^e(k-2) \quad k = 2, 3, 
\ldots, (N_x-4)/2, \\
b^e(k) &= r(2k) - l^e_{k,k-1} b^e(k-1) - l^e_{k,k-2}b(^ek-2)\quad k = 2, 3, 
\ldots, (N_x-4)/2.
\end{align}
In other words, we need the three closest upper diagonals of $U$ plus $a$ and 
$b$. In other words, the total storage requirement is $7N$. Having completed 
the LU decomposition with minimal storage, the solution of the system can be 
obtained very quickly with one forward elimination and a back solve. The back 
solve is very fast since there is only one new addition to the sum for each row
